---
title: "DE on CD5 high vs CD5 low T cells from Matson et al."
output: html_notebook
---

Zoë Steier

Run DE on CD5 high vs low cells (in CD4 and CD8).
Data: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE151395 (one sample: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM4577179)
Paper: https://www.pnas.org/content/pnas/early/2020/06/05/1922525117.full.pdf, https://www.pnas.org/content/117/25/14342 (2020)

Data processing description from GEO: 
The 101 bp paired-end reads were aligned using STAR version 2.5.2a and reads quantified with featureCounts version 1.4.6-p3. Differential expression analysis was performed using DESeq2(71). We used a false discovery rate cutoff of < 1% to define statistical significance.
Genome_build: mm10
Supplementary_files_format_and_content: Raw counts table: 24421 rows x 16 sample columns + geneid column with gene symbols.


# Load required packages
```{r Load pacakges}

# For GEO download
library(Biobase)
library(GEOquery)

# For DESeq2
library(DESeq2)
library(tidyverse)

# For fgsea
library(rlang)
library(stringr)
library(purrr)
library(org.Mm.eg.db)
library(org.Hs.eg.db)
library(biomaRt)
library(stringr)
library(fgsea)
library(parallel)
library(openxlsx)
library(BiocParallel)
register(MulticoreParam(workers = 5))
```


# Download data from GEO

```{r Load data from GEO}
# Only run this once
gset <- getGEOSuppFiles("GSE151395", fetch_files = TRUE) # File to download: GSE151395_raw_counts_genes.txt.gz
getwd() 

# Load the data
geo_filepath <- row.names(gset)[1]

raw <- read_tsv(geo_filepath)
raw
```
# Differential expression analysis with DESeq2

## Load downloaded data
```{r Load the downloaded data}
geo_filepath <- "/data/yosef2/users/zsteier/TCRsignal/TcellSelectionTiming/MatsonPNAS_DE/GSE151395/GSE151395_raw_counts_genes.txt.gz"

raw <- read_tsv(geo_filepath)
raw_mat <- as.matrix(raw[, -1])
row.names(raw_mat) <- raw$geneid

raw_mat[1:5, 1:5]
dim(raw_mat)
```

## Run DESeq2

1. CD4: CD5lo vs CD5hi
2. CD8: CD5lo vs CD5hi

Rank gene lists by padj and logFC.

```{r Make metadata for DESeq2}
meta <- tibble(Sample = colnames(raw)) %>%
  filter(!(Sample == "geneid")) %>%
  mutate(CellType_CD5 = str_replace(Sample, "_[0-9]", "")) %>%
  separate(CellType_CD5, into = c("CellType", "CD5"), sep = "_", remove = FALSE) 
coldata <- as.data.frame(meta)
coldata$CellType <- as.factor(coldata$CellType)
coldata$CellType_CD5 <- as.factor(coldata$CellType_CD5)
coldata$CD5 <- as.factor(coldata$CD5)
row.names(coldata) <- meta$Sample

coldata
```

```{r Run DESeq2}
# Make DESeq2 object
dedsRNASeq <- DESeqDataSetFromMatrix(raw_mat, # un-normalized read counts from raw matrix
                                 colData = coldata,
                                 design = ~ CellType_CD5)

# Run DESeq2
deRNA <- DESeq(dedsRNASeq)

# Extract results
res_CD4 <- results(deRNA, contrast=c("CellType_CD5", "cd4_cd5lo", "cd4_cd5hi"), alpha=0.05)
    
# Make results into tibble
tbDE4 <- as_tibble(res_CD4) %>%
  mutate(Gene = rownames(res_CD4)) %>%
  dplyr::select(Gene, everything()) %>%
  arrange(desc(log2FoldChange))

# View some genes of interest
tbDE4 %>%
  arrange(desc(log2FoldChange)) %>%
  #filter(Gene %in% c("Kcna2", "Tmie", "Cd5"))
  filter(Gene %in% c("Cd5", "Nfkb1", "Nfkbia", "Rela", "Nfkbib", "Ikbkb", "Chuk", "Nfkb2", "Relb"))


# Extract results
res_CD8 <- results(deRNA, contrast=c("CellType_CD5", "cd8_cd5lo", "cd8_cd5hi"), alpha=0.05)
    
# Make results into tibble
tbDE8 <- as_tibble(res_CD8) %>%
  mutate(Gene = rownames(res_CD8)) %>%
  dplyr::select(Gene, everything()) %>%
  arrange(desc(log2FoldChange))

# View genes of interest
tbDE8 %>%
  arrange(desc(log2FoldChange)) %>%
  filter(Gene %in% c("Kcna2", "Tmie", "Cd5", "Cd69", "Nr4a1", "Egr1", "Nfkb1", "Nfatc3"))

sum(is.na(tbDE8$padj))

tbDE8 %>%
  arrange(pvalue)
```
Some p values are NA. See the note "Note on p-values set to NA" on DESeq2: http://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html.
-If only padj is NA, this gene didn't pass the low count filter.
-If LFC is also NA, the baseMean was 0.
-If pvalue is NA, there was an extreme outlier.

Mostly seems due to independent filtering. Could be turned off by setting independentFiltering to FALSE when calling results().


```{r Save results}
write_csv(tbDE8, "DESeq2_Matson_CD8_CD5lo_vs_CD5hi.csv")
write_csv(tbDE4, "DESeq2_Matson_CD4_CD5lo_vs_CD5hi.csv")
```